---
title: "CDF_analysis"
author: "Neo Chatterjee"
date: "2026-06-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r, echo = FALSE}
library(dplyr)
library(lubridate)
library(MASS)


LAG <- data.frame()

#------------likelihood functions

#---typical:
estim <- function(theta, data){
  mu <- theta[1]
  sigma <- theta[2]
  
  p <- pnorm(data$DOY, mu, sigma)
  
  fitted <- p * max(data$count)
  
  return(-sum(dpois(data$count, fitted, log = TRUE)))
}

#---anomalous:
kiyoung <- function(theta, data){
  sigma <- theta[1]
  
  p <- pnorm(data$DOY, mu, sigma)
  
  fitted <- p * max(data$count)
  
  return(-sum(dpois(data$count, fitted, log = TRUE)))
}


#initial guess
start <- c(120, 10)

sites <- c("GREN", "HNRY", "TEA", "BLAN", "SCBI", "SERC", "ORNL", "MLBS", "TREE", "HARV")


ticks <- read.csv("/usr4/ugrad/neochatt/TickForecast/Data/tickLong.csv") |>
  mutate(
    collectDate = as.Date(collectDate),
    year = year(collectDate)
  ) |>
  filter(scientificName == "Ixodes scapularis") |>
  filter(lifeStage == "Nymph") 

phenology <- read.csv("/usr4/ugrad/neochatt/TickForecast/Data/phenology_DOYs.csv")


for(site in sites){
  
  deer <- ticks |>
  filter(siteID == site) |>
  arrange(collectDate) |>
  group_by(collectDate) |>
  summarise(count = sum(processedCount)) 
  deer$year <- year(deer$collectDate)
  deer$DOY <- as.numeric(format(deer$collectDate, "%j"))
  
  deer <- deer[deer$year > 2015, ]
  deer <- na.omit(deer)

  pheno <- phenology |>
  filter(siteID == site)


  years <- min(deer$year):2024
  
  lags <- data.frame(
    siteID = site,
    year = years,
    greenup = vector(mode = "numeric", length = length(years)),
    mid_greenup = vector(mode = "numeric", length = length(years)),
    maturity = vector(mode = "numeric", length = length(years)),
    tick_15 = vector(mode = "numeric", length = length(years)),
    anomalous = vector(mode = "logical", length = length(years))
  )
  
  i <- 1
  
  for(y in years){

  #=============filter to year y
  temp <- deer |>
    filter(year == y)
  
  if(!any(temp$count >= 5)){
      lags$greenup[i] <- NA
      lags$mid_greenup[i] <- NA
      lags$maturity[i] <- NA
      lags$tick_15[i] <- NA
      i <- i+1
      next
    }
  
  if(y == 2020){
    lags$greenup[i] <- NA
    lags$mid_greenup[i] <- NA
    lags$maturity[i] <- NA
    lags$tick_15[i] <- NA
    i <- i+1
    next
  }
  
  
  #=============Locate first peak 
  local_max <- which(
    temp$count > c(-Inf, head(temp$count, -1)) &
      temp$count >= c(tail(temp$count, -1), -Inf)
  )
  first_peak <- local_max[which(temp$count[local_max] >= 5)[1]]
  peak_DOY <- temp$DOY[first_peak]
  
  #=============Find ramp + locate 15th percentile
  ramp <- temp |>
    filter(DOY <= peak_DOY)
  
  if(nrow(ramp) == 1){
    lags$greenup[i] <- NA
    lags$mid_greenup[i] <- NA
    lags$maturity[i] <- NA
    lags$tick_15[i] <- NA
    i <- i+1
    next
  }
  
  if(sum(ramp$count != 0) == 1){
      
      mu <- (ramp$DOY[first_peak-1] + ramp$DOY[first_peak])/2
      
      fit <- optim(par = 2, 
             fn = kiyoung, 
             data = ramp, 
             method = "Brent",
             lower = 0.1,
             upper = 5)
      
      sigma <- fit$par[1]
      print(paste0(site, y, " ", sigma))
      
      c1 <- qnorm(0.15, mean = mu, sd = sigma)
      d1 <- pheno$greenup[pheno$year == y]
      d2 <- pheno$midgreenup[pheno$year == y]
      d3 <- pheno$maturity[pheno$year == y]
      
      lags$greenup[i] <- d1
      lags$mid_greenup[i] <- d2
      lags$maturity[i] <- d3
      lags$tick_15[i] <- c1
      lags$anomalous[i] <- TRUE
      
      i <- i+1
      
      next
    }

  
  fit <- optim(start, fn = estim, data = ramp)

  mu <- fit$par[1]
  sigma <- fit$par[2]
  
  #locate percentile 
  c1 <- qnorm(0.15, mean = mu, sd = sigma)

  
  #===========leaf phenology
  d1 <- pheno$greenup[pheno$year == y]
  d2 <- pheno$midgreenup[pheno$year == y]
  d3 <- pheno$maturity[pheno$year == y]
  
  #==========load into dataframe
  lags$greenup[i] <- d1
  lags$mid_greenup[i] <- d2
  lags$maturity[i] <- d3
  lags$tick_15[i] <- c1
  lags$anomalous[i] <- FALSE
    
  i <- i + 1
  
  }

  LAG <- rbind(LAG, lags)
  
}
  




LAG <- na.omit(LAG)




```



#=========Diagnostic plot (can be used for either species
# or life stage assuming LAG has been generated)

species <- "Ixodes"
lifeStage <- "nymph"

plot(LAG$maturity, LAG$tick_15, pch = 16, cex = 0.7, col = "blue", xlim = c(min(LAG$greenup), max(LAG$maturity)), xlab = "Phenology DOY", ylab = paste0(species, " ", lifeStage, " ", "emergence DOY"))
abline(a =0, b = 1, col = "black", lty = 2)
points(LAG$mid_greenup, LAG$tick_15, pch = 16, cex = 0.7, col = "forestgreen")
points(LAG$greenup, LAG$tick_15, pch = 16, cex = 0.7, col = "red")

legend(
  "topleft",
  legend = c("Greenup", "Mid-greenup", "Maturity"),
  col = c("red", "forestgreen", "blue"),
  pch = c(NA, NA, NA),
  lty = c(1, 1, 1),
  cex = 0.7
)



#============MODEL STATISTICS
summary(lm(LAG$tick_15 ~ LAG$greenup))
summary(lm(LAG$tick_15 ~ LAG$mid_greenup))
summary(lm(LAG$tick_15 ~ LAG$maturity))




#==============cumulative mean DOYs
DOY <- LAG |>
  arrange(siteID, year) |>
  group_by(siteID) |>
  mutate(mean_greenup = cummean(greenup),
         mean_mid = cummean(mid_greenup)) |>
  ungroup()

DOY <- DOY |>
  filter(year > 2016)
  



## AMBLYOMMA
```{r, echo = FALSE}

LAG <- data.frame()

sites <- c("BLAN", "DELA", "KONA", "KONZ", "LENO", "ORNL", "OSBS", "SCBI", "SERC", "TALL", "UKFS")

ticks <- read.csv("/usr4/ugrad/neochatt/TickForecast/Data/tickLong.csv") |>
  mutate(
    collectDate = as.Date(collectDate),
    year = year(collectDate)
  ) |>
  filter(scientificName == "Amblyomma americanum") |>
  filter(lifeStage == "Nymph")

phenology <- read.csv("/usr4/ugrad/neochatt/TickForecast/Data/phenology_DOYs.csv")


for(site in sites){
  
  star <- ticks |>
    filter(siteID == site) |>
    arrange(collectDate) |>
    group_by(collectDate) |>
    summarise(count = sum(processedCount)) 
    star$year <- year(star$collectDate)
    star$DOY <- as.numeric(format(star$collectDate, "%j"))
  
  star <- star[star$year > 2015, ]
  star <- na.omit(star)

  pheno <- phenology |>
  filter(siteID == site)


  years <- min(star$year):2024
  
  lags <- data.frame(
    siteID = site,
    year = years,
    greenup = vector(mode = "numeric", length = length(years)),
    mid_greenup = vector(mode = "numeric", length = length(years)),
    maturity = vector(mode = "numeric", length = length(years)),
    tick_15 = vector(mode = "numeric", length = length(years))
  )
  
  i <- 1
  
  for(y in years){

    #=============filter to year y
    temp <- star |>
      filter(year == y)
    
    if(!any(temp$count >= 5)){
      lags$greenup[i] <- NA
      lags$mid_greenup[i] <- NA
      lags$maturity[i] <- NA
      lags$tick_15[i] <- NA
      i <- i+1
      next
    }
    
    #=============Locate first peak 
    local_max <- which(
      temp$count > c(-Inf, head(temp$count, -1)) &
        temp$count >= c(tail(temp$count, -1), -Inf)
    )
    first_peak <- local_max[which(temp$count[local_max] >= 5)[1]]
    peak_DOY <- temp$DOY[first_peak]
    
    #=============Find ramp + locate 15th percentile
    ramp <- temp |>
      filter(DOY <= peak_DOY) |>
      mutate(
        cum_count = cumsum(count),
        total = sum(count),
        prop = cum_count/total
      )
    
    if(sum(ramp$count != 0) == 1){
      lags$greenup[i] <- NA
      lags$mid_greenup[i] <- NA
      lags$maturity[i] <- NA
      lags$tick_15[i] <- NA
      i <- i+1
      next
    }
    
    fit <- optim(start, fn = estim, data = ramp)

    mu <- fit$par[1]
    sigma <- fit$par[2]
    
    #locate percentile 
    c1 <- qnorm(0.15, mean = mu, sd = sigma)
    
    
    #===========leaf phenology
    d1 <- pheno$greenup[pheno$year == y]
    d2 <- pheno$midgreenup[pheno$year == y]
    d3 <- pheno$maturity[pheno$year == y]
    
    #==========load into dataframe
    lags$greenup[i] <- d1
    lags$mid_greenup[i] <- d2
    lags$maturity[i] <- d3
    lags$tick_15[i] <- c1
      
    i <- i + 1
  
  }

  LAG <- rbind(LAG, lags)
  
}
  


LAG <- na.omit(LAG)

```