---
title: "Exploratory analysis of covariates (LAI, NDVI, EVI) against gross tick counts"
author: "Neo Chatterjee"
date: "2026-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document contains scatterplots of gross tick counts vs each covariate (both time-averaged), with the former variable in turn disaggregated by the following means:

- *All* life stages of *all* species (no disaggregation)
- *Nymphs* of all species 
- *Adults* of all species 
- *Ixodes* *nymphs* only
- *Ixodes* *adults* only
- *Amblyomma* *nymphs* only
- *Amblyomma* *adults* only

As a reminder, the sample size is constrained by the sites (Cary + 10 NEON) for which vegetation data have been downloaded and prepared. 


### ALL-SPECIES PLOTS 
```{r, echo = FALSE}
library(dplyr)
library(lubridate)

ticks <- read.csv("/usr4/ugrad/neochatt/TickForecast/Data/tickLong.csv")
ticks$collectDate <- as.Date(ticks$collectDate)

lai <- read.csv("/usr4/ugrad/neochatt/TickForecast/Data/full_LAIs.csv")
lai$date <- as.Date(lai$date, format = "%m/%d/%Y")

ndvi <- read.csv("/usr4/ugrad/neochatt/TickForecast/Data/full_VIs.csv")
ndvi$date <- as.Date(ndvi$date, format = "%m/%d/%Y")

evi <- read.csv("/usr4/ugrad/neochatt/TickForecast/Data/full_VIs.csv")
evi$date <- as.Date(evi$date, format = "%m/%d/%Y")



ticks <- ticks |>
  mutate(week = floor_date(collectDate, "week"))

lai <- lai |>
  mutate(week = floor_date(date,  "week"))

ndvi <- ndvi |>
  mutate(week = floor_date(date,  "week"))

evi <- evi |>
  mutate(week = floor_date(date,  "week"))

life_stage_groups <- list(
  "all life stages" = c("Larva", "Nymph", "Adult"),
  "nymph" = c("Nymph"),
  "adult" = c("Adult")
)




lai <- lai |>
  group_by(siteID) |>
  summarise(lai = mean(lai_mean, na.rm = TRUE))

ndvi <- ndvi |>
  group_by(siteID) |>
  summarise(ndvi = mean(ndvi_mean, na.rm = TRUE))

evi <- evi |>
  group_by(siteID) |>
  summarise(evi = mean(evi_mean, na.rm = TRUE))




for(stage_name in names(life_stage_groups)){
  
  stages <- life_stage_groups[[stage_name]]
  
  ticks_summary <- ticks |>
    filter(!is.na(collectDate)) |>
    filter(lifeStage %in% stages) |>
    group_by(siteID, collectDate) |>
    summarise(total_count = sum(processedCount, na.rm = TRUE), .groups = "drop") |>
    group_by(siteID) |>
    summarise(mean_count = mean(total_count, na.rm = TRUE), .groups = "drop")
  
  
  #==========LAI=============
  dat <- inner_join(ticks_summary, lai, by = "siteID")
  
  model <- lm(mean_count ~ lai, data = dat)
  
  plot(dat$lai, dat$mean_count,
       pch = 16, cex = 0.7, col = "forestgreen",
       xlab = "LAI",
       ylab = paste0("Mean ", stage_name, " count"),
       main = paste0("Mean ", stage_name, " count vs LAI"))
  
  abline(model, col = "red")
  print(summary(model))
  
  
  #==========NDVI=============
  dat <- inner_join(ticks_summary, ndvi, by = "siteID")
  
  model <- lm(mean_count ~ ndvi, data = dat)
  
  plot(dat$ndvi, dat$mean_count,
       pch = 16, cex = 0.7, col = "forestgreen",
       xlab = "NDVI",
       ylab = paste0("Mean ", stage_name, " count"),
       main = paste0("Mean ", stage_name, " count vs NDVI"))
  
  abline(model, col = "red")
  print(summary(model))
  
  
  #==========EVI=============
  dat <- inner_join(ticks_summary, evi, by = "siteID")
  
  model <- lm(mean_count ~ evi, data = dat)
  
  plot(dat$evi, dat$mean_count,
       pch = 16, cex = 0.7, col = "forestgreen",
       xlab = "EVI",
       ylab = paste0("Mean ", stage_name, " count"),
       main = paste0("Mean ", stage_name, " count vs EVI"))
  
  abline(model, col = "red")
  print(summary(model))
  

}

```


### IXODES ONLY

```{r, echo = FALSE}
library(dplyr)
library(lubridate)

ticks <- read.csv("/usr4/ugrad/neochatt/TickForecast/Data/tickLong.csv")
ticks$collectDate <- as.Date(ticks$collectDate)

lai <- read.csv("/usr4/ugrad/neochatt/TickForecast/Data/full_LAIs.csv")
lai$date <- as.Date(lai$date, format = "%m/%d/%Y")

ndvi <- read.csv("/usr4/ugrad/neochatt/TickForecast/Data/full_VIs.csv")
ndvi$date <- as.Date(ndvi$date, format = "%m/%d/%Y")

evi <- read.csv("/usr4/ugrad/neochatt/TickForecast/Data/full_VIs.csv")
evi$date <- as.Date(evi$date, format = "%m/%d/%Y")



ixodes <- ticks |>
  mutate(week = floor_date(collectDate, "week")) |>
  filter(scientificName %in% c("Ixodes scapularis"))

amblyomma <- ticks |>
  mutate(week = floor_date(collectDate, "week")) |>
  filter(scientificName %in% c("Amblyomma americanum"))


lai <- lai |>
  mutate(week = floor_date(date,  "week"))

ndvi <- ndvi |>
  mutate(week = floor_date(date,  "week"))

evi <- evi |>
  mutate(week = floor_date(date,  "week"))

life_stage_groups <- list(
  "all life stages" = c("Larva", "Nymph", "Adult"),
  "nymph" = c("Nymph"),
  "adult" = c("Adult")
)




lai <- lai |>
  group_by(siteID) |>
  summarise(lai = mean(lai_mean, na.rm = TRUE))

ndvi <- ndvi |>
  group_by(siteID) |>
  summarise(ndvi = mean(ndvi_mean, na.rm = TRUE))

evi <- evi |>
  group_by(siteID) |>
  summarise(evi = mean(evi_mean, na.rm = TRUE))





for(stage_name in names(life_stage_groups)){
  
  stages <- life_stage_groups[[stage_name]]
  
  ticks_summary <- ixodes |>
    filter(!is.na(collectDate)) |>
    filter(lifeStage %in% stages) |>
    group_by(siteID, collectDate) |>
    summarise(total_count = sum(processedCount, na.rm = TRUE), .groups = "drop") |>
    group_by(siteID) |>
    summarise(mean_count = mean(total_count, na.rm = TRUE), .groups = "drop")
  
  
  #==========LAI=============
  dat <- inner_join(ticks_summary, lai, by = "siteID")
  
  model <- lm(mean_count ~ lai, data = dat)
  
  plot(dat$lai, dat$mean_count,
       pch = 16, cex = 0.7, col = "forestgreen",
       xlab = "LAI",
       ylab = paste0("Mean Ixodes ", stage_name, " count"),
       main = paste0("Mean Ixodes ", stage_name, " count vs LAI"))
  
  abline(model, col = "red")
  print(summary(model))
  
  
  #==========NDVI=============
  dat <- inner_join(ticks_summary, ndvi, by = "siteID")
  
  model <- lm(mean_count ~ ndvi, data = dat)
  
  plot(dat$ndvi, dat$mean_count,
       pch = 16, cex = 0.7, col = "forestgreen",
       xlab = "NDVI",
       ylab = paste0("Mean Ixodes ", stage_name, " count"),
       main = paste0("Mean Ixodes ", stage_name, " count vs NDVI"))
  
  abline(model, col = "red")
  print(summary(model))
  
  
  #==========EVI=============
  dat <- inner_join(ticks_summary, evi, by = "siteID")
  
  model <- lm(mean_count ~ evi, data = dat)
  
  plot(dat$evi, dat$mean_count,
       pch = 16, cex = 0.7, col = "forestgreen",
       xlab = "EVI",
       ylab = paste0("Mean Ixodes ", stage_name, " count"),
       main = paste0("Mean Ixodes ", stage_name, " count vs EVI"))
  
  abline(model, col = "red")
  print(summary(model))
  
  
  
}

```





### AMBLYOMMA ONLY
```{r, echo = FALSE}

for(stage_name in names(life_stage_groups)){
  
  stages <- life_stage_groups[[stage_name]]
  
  ticks_summary <- amblyomma |>
    filter(!is.na(collectDate)) |>
    filter(lifeStage %in% stages) |>
    group_by(siteID, collectDate) |>
    summarise(total_count = sum(processedCount, na.rm = TRUE), .groups = "drop") |>
    group_by(siteID) |>
    summarise(mean_count = mean(total_count, na.rm = TRUE), .groups = "drop")
  
  
  #==========LAI=============
  dat <- inner_join(ticks_summary, lai, by = "siteID")
  
  model <- lm(mean_count ~ lai, data = dat)
  
  plot(dat$lai, dat$mean_count,
       pch = 16, cex = 0.7, col = "forestgreen",
       xlab = "LAI",
       ylab = paste0("Mean Amblyomma ", stage_name, " count"),
       main = paste0("Mean Amblyomma ", stage_name, " count vs LAI"))
  
  abline(model, col = "red")
  print(summary(model))
  
  
  #==========NDVI=============
  dat <- inner_join(ticks_summary, ndvi, by = "siteID")
  
  model <- lm(mean_count ~ ndvi, data = dat)
  
  plot(dat$ndvi, dat$mean_count,
       pch = 16, cex = 0.7, col = "forestgreen",
       xlab = "NDVI",
       ylab = paste0("Mean Amblyomma ", stage_name, " count"),
       main = paste0("Mean Amblyomma ", stage_name, " count vs NDVI"))
  
  abline(model, col = "red")
  print(summary(model))
  
  
  #==========EVI=============
  dat <- inner_join(ticks_summary, evi, by = "siteID")
  
  model <- lm(mean_count ~ evi, data = dat)
  
  plot(dat$evi, dat$mean_count,
       pch = 16, cex = 0.7, col = "forestgreen",
       xlab = "EVI",
       ylab = paste0("Mean Amblyomma ", stage_name, " count"),
       main = paste0("Mean Amblyomma ", stage_name, " count vs EVI"))
  
  abline(model, col = "red")
  print(summary(model))
  
  
  
}


```








