---
title: "Land cover analysis (NLCD, fragstats)"
author: "Neo Chatterjee"
date: "2026-05-13"
output: html_document
---

This document contains exploratory analyses of three remotely sensed variables, (a): the % of a site composed by its dominant land cover, (b): habitat edge, and (c): the dominant land cover per se. Said analyses can in turn be split into two halves:

- (1): Associations between either of the first two covariates and *gross* tick counts (the latter disaggregated by species and nymph or adult status, thus yielding four plots in all for each covariate)
- (2): The box-and-whisker distribution of the most recently estimated parameter values for sites sharing a dominant land cover 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



### GROSS COUNTS (%-dominant land cover)
```{r, echo = FALSE}
library(dplyr) 

NLCD <- read.csv("/usr4/ugrad/neochatt/TickForecast/Data/full_NLCD.csv")
frag <- read.csv("/usr4/ugrad/neochatt/TickForecast/Data/fragstats.csv")

dat <- inner_join(NLCD, frag, by = "siteID")

#Ixodes nymphs
mod <- lm(mean_ixodes_nymphs ~ pct_dominant, data = dat)

plot(dat$pct_dominant, dat$mean_ixodes_nymphs, 
     xlab = "% dominant", 
     ylab = "Mean Ixodes nymph count", 
     main = "Mean Ixodes nymphs vs % dominant land cover", 
     pch = 16, cex = 0.7)
abline(mod, col = "red")
print(summary(mod))

#Ixodes adults
mod <- lm(mean_ixodes_adults ~ pct_dominant, data = dat)

plot(dat$pct_dominant, dat$mean_ixodes_adults, 
     xlab = "% dominant", 
     ylab = "Mean Ixodes adult count", 
     main = "Mean Ixodes adults vs % dominant land cover", 
     pch = 16, cex = 0.7)
abline(mod, col = "red")
print(summary(mod))






#Amblyomma nymphs
mod <- lm(mean_amblyomma_nymphs ~ pct_dominant, data = dat)

plot(dat$pct_dominant, dat$mean_amblyomma_nymphs,
     xlab = "% dominant",
     ylab = "Mean Amblyomma nymph count",
     main = "Mean Amblyomma nymphs vs % dominant land cover",
     pch = 16, cex = 0.7)
abline(mod, col = "red")
print(summary(mod))

#Amblyomma adults
mod <- lm(mean_amblyomma_adults ~ pct_dominant, data = dat)

plot(dat$pct_dominant, dat$mean_amblyomma_adults,
     xlab = "% dominant",
     ylab = "Mean Amblyomma adult count",
     main = "Mean Amblyomma adults vs % dominant land cover",
     pch = 16, cex = 0.7)
abline(mod, col = "red")
print(summary(mod))




```


### GROSS COUNTS (edge)
```{r, echo = FALSE}


# Ixodes nymphs
mod <- lm(mean_ixodes_nymphs ~ edge_m_per_ha, data = dat)

plot(dat$edge_m_per_ha, dat$mean_ixodes_nymphs,
     xlab = "Edge meters per hectare",
     ylab = "Mean Ixodes nymph count",
     main = "Mean Ixodes nymphs vs edge density",
     pch = 16, cex = 0.7)

abline(mod, col = "red")

print(summary(mod))


# Ixodes adults
mod <- lm(mean_ixodes_adults ~ edge_m_per_ha, data = dat)

plot(dat$edge_m_per_ha, dat$mean_ixodes_adults,
     xlab = "Edge meters per hectare",
     ylab = "Mean Ixodes adult count",
     main = "Mean Ixodes adults vs edge density",
     pch = 16, cex = 0.7)

abline(mod, col = "red")

print(summary(mod))


# Amblyomma nymphs
mod <- lm(mean_amblyomma_nymphs ~ edge_m_per_ha, data = dat)

plot(dat$edge_m_per_ha, dat$mean_amblyomma_nymphs,
     xlab = "Edge meters per hectare",
     ylab = "Mean Amblyomma nymph count",
     main = "Mean Amblyomma nymphs vs edge density",
     pch = 16, cex = 0.7)

abline(mod, col = "red")

print(summary(mod))


# Amblyomma adults
mod <- lm(mean_amblyomma_adults ~ edge_m_per_ha, data = dat)

plot(dat$edge_m_per_ha, dat$mean_amblyomma_adults,
     xlab = "Edge meters per hectare",
     ylab = "Mean Amblyomma adult count",
     main = "Mean Amblyomma adults vs edge density",
     pch = 16, cex = 0.7)

abline(mod, col = "red")

print(summary(mod))


```






### PARAMETERS (IXODES)
```{r, echo = FALSE}
library(ggplot2)


library(dplyr)
library(lubridate)
library(broom)

#obtain parameter .csv's
files <- list.files("/usr4/ugrad/neochatt/TickForecast/Data", pattern = "Ixodes", full.names = TRUE)

#parameter names
pars <- c("phi.l.mu", "phi.n.mu", "phi.a.mu", "theta.ln", "theta.na", "beta[1]", "beta[2]", "beta[3]", "beta[4]", "beta[5]", "beta[6]", "beta[7]", "beta[8]", "beta[9]", "beta[10]", "beta[11]", "beta[12]", "beta[13]", "beta[14]")

for(par in pars){
  
  deer_files <- list()

  
  for (f in files) {
    dat <- read.csv(f)
    dat <- dat |> filter(node == par)
    dat$start.date <- as.Date(dat$start.date)
    deer_files[[length(deer_files) + 1]] <- dat
  }

  deer_files <- bind_rows(deer_files)

  deer_files <- deer_files |>
    mutate(week = floor_date(start.date, "week")) |>
    group_by(week, siteID) |>
    mutate(par = mean(mean, na.rm = TRUE))
  
  params <- deer_files |>
    group_by(siteID) |>
    summarize(param_value = mean[which.max(as.Date(start.date))])
  
  
  kat <- frag |>
    inner_join(params, by = "siteID")
  
  kat <- kat |>
    inner_join(NLCD, by = "siteID")
  
  
  
  p <- ggplot(kat, 
              aes(x = factor(lc_dominant), 
                  y = param_value,
                  fill = factor(lc_dominant))) +
    geom_boxplot() +
    labs(
      title = par,
      x = "Dominant land cover",
      y = "Parameter value",
      fill = "Dominant land cover"
    ) + 
    scale_fill_discrete(labels = c(
      deciduous_forest_pct = "Deciduous forest",
      mixed_forest_pct = "Mixed forest",
      woody_wetland_pct = "Woody wetland"))

  

  
  print(p)
  
}



```

### PARAMETERS (AMBLYOMMA)
```{r, echo = FALSE}

library(ggplot2)


library(dplyr)
library(lubridate)
library(broom)

#obtain parameter .csv's
files <- list.files("/usr4/ugrad/neochatt/TickForecast/Data", pattern = "Amblyomma", full.names = TRUE)

#parameter names
pars <- c("phi.l.mu", "phi.n.mu", "phi.a.mu", "theta.ln", "theta.na", "beta[1]", "beta[2]", "beta[3]", "beta[4]", "beta[5]", "beta[6]", "beta[7]", "beta[8]", "beta[9]", "beta[10]", "beta[11]", "beta[12]", "beta[13]", "beta[14]")

for(par in pars){
  
  deer_files <- list()

  
  for (f in files) {
    dat <- read.csv(f)
    dat <- dat |> filter(node == par)
    dat$start.date <- as.Date(dat$start.date)
    deer_files[[length(deer_files) + 1]] <- dat
  }

  deer_files <- bind_rows(deer_files)

  deer_files <- deer_files |>
    mutate(week = floor_date(start.date, "week")) |>
    group_by(week, siteID) |>
    mutate(par = mean(mean, na.rm = TRUE))
  
  params <- deer_files |>
    group_by(siteID) |>
    summarize(param_value = mean[which.max(as.Date(start.date))])
  
  
  kat <- frag |>
    inner_join(params, by = "siteID")
  
  kat <- kat |>
    inner_join(NLCD, by = "siteID")
  
  
  
  p <- ggplot(kat, 
              aes(x = factor(lc_dominant), 
                  y = param_value,
                  fill = factor(lc_dominant))) +
    geom_boxplot() +
    labs(
      title = par,
      x = "Dominant land cover",
      y = "Parameter value",
      fill = "Dominant land cover"
    ) +
    scale_fill_discrete(labels = c(
      deciduous_forest_pct = "Deciduous forest",
      evergreen_forest_pct = "Evergreen forest",
      grassland_pct = "Grassland",
      woody_wetland_pct = "Woody wetland"))

  

  
  print(p)
  
}





```







