---
title: "off-switch"
author: "Neo Chatterjee"
date: "2026-08-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE}
library(dplyr)
library(lubridate)
library(MASS)


LAG <- data.frame()



sites <- c("GREN", "HNRY", "TEA", "BLAN", "SCBI", "SERC", "ORNL", "MLBS", "TREE", "HARV")


years <- 2016:2024


ticks <- read.csv("/usr4/ugrad/neochatt/TickForecast/Data/tickLong.csv") |>
  mutate(
    collectDate = as.Date(collectDate),
    year = year(collectDate)
  ) |>
  filter(scientificName == "Ixodes scapularis") |>
  filter(lifeStage == "Nymph") 




for(site in sites){
  for(y in years){
    
    deer <- ticks |> filter(siteID == site, year == y) |> 
                     arrange(collectDate) |> 
                     group_by(collectDate)|> 
                     summarise(count = sum(processedCount)) 
    
    deer$year <- year(deer$collectDate) 
    deer$DOY <- as.numeric(format(deer$collectDate, "%j")) 
    
    deer <- deer |> arrange(DOY) |> 
                    mutate(cum_prop = cumsum(count) / sum(count, na.rm = TRUE)) 
    
    tick_85 <- deer$DOY[which(deer$cum_prop >= 0.85)[1]]
  
    lag <- data.frame(siteID = site,
                      year = y,
                      tick_85 = tick_85)
    
    LAG <- rbind(LAG, lag)
    
  }
}


LAG <- na.omit(LAG)


```


#=========Diagnostic plot (can be used for either species
# or life stage assuming LAG has been generated)

species <- "Ixodes"
lifeStage <- "nymph"

plot(LAG$maturity, LAG$tick_85, pch = 16, cex = 0.7, col = "blue", xlim = c(min(LAG$greenup), max(LAG$maturity)), xlab = "Phenology DOY", ylab = paste0(species, " ", lifeStage, " ", "emergence DOY"))
abline(a =0, b = 1, col = "black", lty = 2)
points(LAG$mid_greenup, LAG$tick_85, pch = 16, cex = 0.7, col = "forestgreen")
points(LAG$greenup, LAG$tick_85, pch = 16, cex = 0.7, col = "red")

legend(
  "topleft",
  legend = c("Greenup", "Mid-greenup", "Maturity"),
  col = c("red", "forestgreen", "blue"),
  pch = c(NA, NA, NA),
  lty = c(1, 1, 1),
  cex = 0.7
)



#============MODEL STATISTICS
summary(lm(LAG$tick_15 ~ LAG$greenup))
summary(lm(LAG$tick_15 ~ LAG$mid_greenup))
summary(lm(LAG$tick_15 ~ LAG$maturity))



